A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-12-04 analysis/paper/paper.Rmd
Writing through supplementary

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating palimpsests of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1. A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```

>>>>

Following the end of the Weichselian and the final retreat of the Fennoscandian Ice Sheet [e.g. @hughes2016; @stroeven2016], the isostatic rebound has been so severe that most areas of Norway have been subject to a continuous relative sea-level regression, despite corresponding eustatic sea-level rise [e.g. @lambeck1998; @svendsen1987]. In other words, the RSL has been dropping throughout prehistory. As this process is the result of glacial loading, the rate of uplift is more severe towards the centre of the ice sheet. Thus, some areas on the outer coast have had a more stable RSL or been subject to marine transgression [e.g. @romundset2015; @svendsen1987]. These conditions are directly reflected in the archaeological record. In areas where the sea-level has been stable over longer periods of time, people have often reused the coastal site locations multiple times and over long time-spans, creating a mix of settlement events that are difficult to disentangle [@reitan2009]. Transgression phases, on the other hand, can lead to complete destruction of the sites, or bury them in marine sediments, leading to a hiatus in the archaeological record for certain sub-phases in the impacted areas [@bjerck2008]. Comparatively, given a continuous and still ongoing shoreline regression from as high as c. 220 m above present sea-level in the inner Oslo fjord, any one location in south-eastern Norway has only been shore-bound within a relatively limited time-span, and the sites have not been impacted by any transgressions [@hafsten1957; @hafsten1983; @sørensen1979]. This makes the region especially useful for evaluating the assumption of a shore-bound settlement over a long and continuous time-span.

@@

```{r example-1, echo = FALSE, fig.cap = "Example site Hegna vest 1 (Fossum 2017). A) Location of the site in the present day landscape. The red outline is the site limit. B) Radiocarbon dates associated with the site. Fill colour indicates what dates are assumed to belong to the same settlement phase. Multiple dates are modelled using the Boundary function in OxCal and then summed. The red outline indicates that the date does not match the typological indicators in the artefact assemblage of the site. C) The location of the site within the study area relative to isobases of the employed displacement curves. D) Displacement curve interpolated to the site location based on the distance between the site and the isobases of the two closest displacement curves.", out.width = "500px", fig.align="center"}
knitr::include_graphics(here("analysis/figures/example_map.png"))
```
